---
title: "Updating White Racial Attitude Change"
output: html_document
date: "`r Sys.Date()`"
urlcolor: blue
header-includes: 
## add latex packages here if wanted 
  - \usepackage{hayesmacros} # Macros file from origin
  - \usepackage[default]{sourcesanspro}
  - \usepackage[T1]{fontenc}
  - \usepackage{siunitx} # needed to render something
  - \usepackage{placeins}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
mainfont: SourceSansPro
linestretch: 1 # line spacing for document

---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  tidy = TRUE, # clean up code
  tidy.opts=list(width.cutoff=50), ## wrap text width for tidying
  echo = FALSE,          # show code
  warning = FALSE,              # don't show warnings
  message = FALSE,              # don't show messages (less serious warnings)
  cache = TRUE,                # set to TRUE to save results from last compilation, otherwise re-runs all code even if it hasn't changed
  fig.align = "center",         # center figures
  #attr.source = '.numberLines',  # add line numbers to code
  dev = "cairo_pdf" # changing graphical device for font
)
```

```{r packages}
library(ggplot2)
library(marginaleffects)
library(patchwork)
```

```{r models}
source("~/Dropbox/Data/ANES/2024/anes24_riclpm_rr.R")
source("~/Dropbox/Data/ANES/2024/anes24_riclpm_ft.R")
```


## The Pattern and Problem

Evidence: BJPS, VOTER publications, Jardina and Ollerenshaw[^1]

[^1]: These include assessments of snapshots of the U.S. public [since the 1980s](https://academic.oup.com/poq/article/86/S1/576/6617224) or work tracking the same respondents [between 2011 and 2024](https://democracyfund.org/idea/pushed-and-pulled/)

This has consequences. 

**Takeaways**

* White racial attitude polarization since 2016 isn't merely party switching, it's attitude change.
* Scholars should be careful 1) testing hypothesis using cross-sectional data and 2) interpreting associations between racial attitudes or partisanship and outcomes of interest.

```{r cars}
summary(cars)
```

## Racial Attitude Trends: 2016-2024

```{r means}
p_means_rr + p_means_wb
```

## Updating Bidirectionality Test with New Data and Method

