---
title: "Updating White Racial Attitude Change"
output: html_document
date: "`r Sys.Date()`"
urlcolor: blue
header-includes: 
## add latex packages here if wanted 
  - \usepackage{hayesmacros} # Macros file from origin
  - \usepackage[default]{sourcesanspro}
  - \usepackage[T1]{fontenc}
  - \usepackage{siunitx} # needed to render something
  - \usepackage{placeins}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
mainfont: SourceSansPro
linestretch: 1 # line spacing for document

---

```{r options, include = FALSE}
knitr::opts_chunk$set(
  tidy = TRUE, # clean up code
  #tidy.opts=list(width.cutoff=50), ## wrap text width for tidying
  echo = FALSE,          # show code
  warning = FALSE,              # don't show warnings
  message = FALSE,              # don't show messages (less serious warnings)
  cache = TRUE,                # set to TRUE to save results from last compilation, otherwise re-runs all code even if it hasn't changed
  fig.align = "center"         # center figures
  #attr.source = '.numberLines',  # add line numbers to code
  # dev = "cairo_pdf" # changing graphical device for font
)
```

```{r packages}
library(ggplot2)
library(marginaleffects)
library(patchwork)
```

```{r models}
source("~/Dropbox/Data/ANES/2024/anes24_riclpm_rr.R")
source("~/Dropbox/Data/ANES/2024/anes24_riclpm_ft.R")
```

```{r mean_plots}
p_rr <- ggplot(subset(rr_means, group == "D.16 D.20 D.24" | group == "R.16 R.20 R.24"), 
       aes(x = year, y = value, group = group, color = group)) +
  geom_point(size = 2) +
  geom_line() +
  theme_bw() +
  # facet_wrap(~group, nrow = 9) +
  labs(x = "", y = "", title = "Explanations for Black Americans' Status") +
  # labs(x = "", y = "", title = "White Racial Resentment by Party Flow",
  #      caption = "Analyses weighted.") +
  scale_color_manual(name = "", values = c("#00AEF3", "#E81B23")) +
  annotate("text", x = 1.25, y = .27, color = "#00AEF3", label = "Democrats") +
  annotate("text", x = 1.25, y = .75, color = "#E81B23", label = "Republicans") +
  # scale_shape_manual(name = "", values = c(1, 16)) +
  scale_size(guide = F) +
  scale_y_continuous(breaks = c(0, 1), limits = c(0, 1),
                     labels = c("Structural\nBarriers", "Individual\nEffort")) +
  # scale_y_continuous(limits = c(0, 1)) +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 12),
        plot.title = element_text(size = 14),
        strip.text = element_text(size = 10),
        legend.position = "none",
        strip.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

p_wb <- ggplot(subset(wb_means, group == "D.16 D.20 D.24" | group == "R.16 R.20 R.24"), 
       aes(x = year, y = value, group = group, color = group)) +
  geom_point(size = 2) +
  geom_line() +
  theme_bw() +
  # facet_wrap(~group, nrow = 9) +
  labs(x = "", y = "", title = "Relative Favorability of Black Americans") +
  scale_color_manual(name = "", values = c("#00AEF3", "#E81B23")) +
  # scale_shape_manual(name = "", values = c(1, 16)) +
  scale_size(guide = F) +
  scale_y_continuous(breaks = c(0, .5, 1), limits = c(0, 1),
                     labels = c("Pro\nBlack", "Neutral", "Pro\nWhite")) +
  # scale_y_continuous(limits = c(0, 1)) +
  theme(axis.text = element_text(size = 12),
        axis.title = element_text(size = 12),
        plot.title = element_text(size = 14),
        strip.text = element_text(size = 10),
        legend.position = "none",
        strip.background = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

**In Brief**

* White racial attitude polarization since 2016 remains a mix of party switching and attitude change.
* Scholars should be careful 1) testing hypotheses using cross-sectional data and 2) interpreting associations between racial attitudes or partisanship and outcomes of interest.


It is commonplace to identify an increasingly tight connection between the views white Americans hold about Black people and their partisanship. These trends appear in snapshots of the U.S. public [since the 1980s](https://academic.oup.com/poq/article/86/S1/576/6617224) or work tracking the same respondents [between 2011 and 2024](https://democracyfund.org/idea/pushed-and-pulled/). While the conventional view is this comes from people switching parties, I've found [evidence](https://www.amengelhardt.com/files/engelhardt_PID_racial_attitudes_bjps.pdf) that since 2008, this conventional view is complemented by one where whites are actually changing their attitudes. These shifts [appear sincere](https://www.amengelhardt.com/files/engelhardt_attitude_change_ajps.pdf), and have paid off in white Democrats increasingly favoring [Black politicians](https://osf.io/preprints/osf/bjhz2) in part out of a desire to [address racial inequality](https://www.cambridge.org/core/journals/american-political-science-review/article/white-democrats-growing-support-for-black-politicians-in-the-era-of-the-great-awokening/CD6861093C88839E6955F4BCC5BEA339).

But the argument I advanced hinged on background social and political context, with the evidence accumulated ending in 2016. And much has changed since then. So, with the release of the 2024 [American National Election Study](https://electionstudies.org/data-center/2024-time-series-study/), I found it useful to update my analysis with these new data. I also apply a statistical method more appropriate for identifying attitudinal shifts [within individuals](http://doi.apa.org/getdoi.cfm?doi=10.1037/a0038889).

While preliminary until the ANES finalizes the 2024 release, I find a continued bidirectional relationship between whites' views of Black Americans and their partisanship. This is clearest on a racial attitude measure that divides respondents according to how much they attribute Black Americans' social and economic status to individual effort or systemic discrimination.[^1] This has consequences.[^2]

[^1]: This measure, known as [racial resentment](https://press.uchicago.edu/ucp/books/book/chicago/D/bo3620441.html), has a [contentious history](https://press.uchicago.edu/ucp/books/book/chicago/R/bo3617330.html). But I use it to take seriously its intention to capture the ["political role of whites' racial attitudes"](https://psycnet.apa.org/record/1988-97566-004), a view which can account for its clearer connections with partisanship than other manifestations fo racial attitude.
[^2]: Dan Hopkins's [*Stable Condition*](https://www.russellsage.org/publications/stable-condition) offers a nice explanation for this inferential challenge.


## Racial Attitude Trends: 2016-2024

```{r means, out.height="100%", out.width = "100%", fig.height=5, fig.width= 12, fig.cap="Trends in Racial Attitudes among Consistent Partisans."}
(p_rr + p_wb) + plot_annotation(tag_levels = 'A')
```

## Updating Bidirectionality Test with New Data and Method


```{r riclpm, out.height="100%", out.width = "100%", fig.height=8, fig.width= 12, fig.cap="Individual-level change in Partisanship and Racial Attitudes. Results from random intercepts cross-lagged panel model."}
(plot_rr / plot_wb ) + plot_annotation(tag_levels = 'A') 
```
